The Ky Fan Norms and Beyond: Dual Norms and Combinations for Matrix Optimization

### 2025-11-30 article_final/main.tex
Added Medium logs + minor changes in the article

We then compare the performance of these algorithm families on various model and real-world problems:
\begin{itemize}
	\item Synthetic least squares experiment \sectionref{subsec:lls_exps}
	\item CIFAR-10 airbench \citep{cifar2023airbench}
	\item Pre-training NanoGPT on FineWeb dataset \citep{modded_nanogpt_2024}
	\item Fine-tuning NanoGPT on Tiny Stories \citep{eldan2023tinystoriessmalllanguagemodels}
\end{itemize}

@@

Training a neural network is essentially an optimization of a function of several weight matrices and a few vectors. Let us start by considering the problem of minimizing a differentiable function of a \emph{single} matrix:
\begin{equation}\label{eq:matrix_problem}
	F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,
\end{equation}
We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which does not have to coincide with the Frobenius norm $\normf{\cdot} = \<\cdot,\cdot>$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ that is associated with $\norm{\cdot}$ is defined as
\begin{equation}\label{eq:dual_norm}
	\norm{\mG}^\dagger = \sup_{\mD \in \Rmn\,:\norm{\mD}\leq 1} \<\mG,\mD>\,.
\end{equation}

Such problems can be solved with an iterative algorithm based on the Linear Minimization Oracle (LMO):
\begin{equation}\label{eq:lmo}
	\mathrm{LMO}(\mM^t) \in \argmin_{\mD \in \cS} \<\mM^t, \mD>\,,
\end{equation}
where $\mM^t$ is the gradient (or momentum) of $F$ at $\mX^t$ and $\cS \subset \Rmn$ is some set. The update of the algorithm is defined as follows:
\begin{equation}\label{eq:simple_update}
	\mX^{t + 1} = \mX^t + \gamma_t\mathrm{LMO}\left(\mM^t\right)\,.
\end{equation}

We are particularly interested in the case when $\cS$ is a unit ball in some norm $\norm{\cdot}$:
\[
\cS = \cB_{\norm{\cdot}} = \{ \mD \in \Rmn \mid \norm{\mD} \leq 1 \}\,.
\]
In this case,
\[
\argmin_{\mD \in \cS} \<\mM^t, \mD> = - \{ \mD \in \cB_{\norm{\cdot}} \mid \<\mM^t, \mD> = \norm{\mM^t}^\dagger\}\,,
\]
and the update for $\mX^t$ in \equationref{eq:simple_update} simplifies to
\begin{equation}\label{eq:our_update}
	\mX^{t+1} = \mX^{t} - \gamma_t \{\mD \in \cB_{\norm{\cdot}} \mid \<\mM^t, \mD> = \norm{\mM^t}^\dagger\}\,.
\end{equation}

@@

\begin{corollary}\label{corollary:lin_comb_alg}
	Let there be a finite family of LMO based algorithms indexed by $i = 1,\dots,n$, where the update of the $i$-th algorithm is defined by
	\[
	\mX^{t + 1} - \mX^{t} = \gamma_t \mathrm{LMO}_{i}(\mM^{t})\,,
	\]
	and $\mathrm{LMO}_i$ corresponds to the unit ball of norm $\norm{\cdot}_i$. For arbitrary non-negative $\alpha_1,\dots,\alpha_n$, the algorithm with the update given by
	\[
	\mX^{t + 1} - \mX^{t} = \gamma_t \sum_{i = 1}^n \alpha_i \mathrm{LMO}_{\norm{\cdot}_{(i)}}(\mM^t)
	\]
	is an LMO-algorithm itself, with LMO corresponding to the unit ball of the norm $\norm{\cdot}$ dual to the norm given by $\sum_{i = 1}^n \alpha_i \norm{\cdot}_{(i)}^\dagger$.
\end{corollary}

@@

Similarly, combining Fanion-$k$ with SignSGD yields \emph{S-Fanion-$k$} with update
\begin{equation}\label{eq:sign_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\eta\sign(\mM^t)\right)\,,
\end{equation}
where $\eta$ is a scaling coefficient specific to SignSGD.

@@

First, we compare F-Fanions on the following convex $L$-smooth problem:
\begin{equation}\label{eq:lls}
  F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN> \to \min_{\mX \in \Rmn}\,
\end{equation}
where $\mX \in \Rmn$, $m=500$ and $n=500$ are typical dimensions of a neural network weight matrix, $\mS \in \Rmn$, $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. The spectra of $\mM$ and $\mN$ are uniformly distributed in the $(0, 1)$ interval. We set $\mS=0$, and $\mX^0 \sim 0.1 \cN(0, \mI)$.

We run different Fanions and their respective F-Fanions with $\alpha=1/2$: Neon (Fanion-1), Fanion-2, Fanion-10, Fanion-100, and Muon (Fanion-500). Also, we test S-Fanions with $\alpha=1/2$ and \texttt{sign\_lr\_coeff=0.01}. We test them against Normalized SGD and SignSGD, which are also F-Fanion and S-Fanion respectively with $\alpha = 0$ and an arbitrary $k$.

Since theoretical bounds \citep{kovalev2025understanding, riabinin2025gluon} rely on a very loose norm bound $\norm{\cdot}\le \rho \normf{\cdot}$, we do not derive learning rate or Nesterov momentum from the smoothness constants that also depend on the norm choice. Rather, we find such \texttt{(lr, momentum)} pair that the loss threshold $0.001$ is reached in a minimal number of iterations. This setting is the most realistic, and at the same time aligns with the corollaries of the convergence theorems where constant learning rate and momentum coefficients are proposed.

We perform a grid search for \texttt{momentum} $\in \{0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95\}$ and for \texttt{lr}. For Muon's, F-Muon's, S-Muon's \texttt{lr} $\in [0.005, 0.0020]$. For SignSGD, \texttt{lr} $\in 0.01 \times [0.005, 0.020]$, and for NSGD \texttt{lr} $\in [0.01, 0.10]$. The tuned parameters and the number of iterations to converge to $0.001$ are in \tableref{tbl:lls_lrs}. \texttt{lr} and \texttt{momentum}, and \texttt{sign\_lr\_coeff} for Fanions, F-Fanions, and S-Fanions are the same as for Muon, F-Muon, and S-Muon respectively due to the impossibility to tune them properly: their losses decrease too slow to reach 0.001.

The results are presented in \figureref{fig:lls} and detailed in \sectionref{sec:lls_plots_section}.

Both F-Muon and S-Muon converge faster and to a lower loss and a lower Frobenius norm of the gradient than NSGD, Muon, or SignSGD.

@@

@@

@@

We compare the algorithms on the CIFAR-10 airbench \citep{cifar2023airbench}. First, to test the impact of $\alpha$, we run F-Muon for different $\alpha$ with the same \texttt{lr=0.24(1 - step/total\_steps), momentum=0.65, nesterov=True} and weight normalization, as have been finetuned by Keller Jordan, 10 repetitions for each $\alpha$. We record the accuracy after 8 epochs of training (\figureref{fig:muon_alphas}).

Then we tune F-Muon with $\alpha = 0.5$. Tuned parameters are \texttt{lr=0.4(1 - step/total\_steps), momentum=0.6, nesterov = True}. Tuned F-Muon reaches $94.02 \pm 0.13\%$ after 8 epochs (averaged by 200 iterations), matching Muon's accuracy and variance. Once again we measure \texttt{val\_accuracy}($\alpha$) curve (\figureref{fig:fmuon_alphas}). We notice that even when $\alpha=0.1$, the accuracy is much higher than in the case of vanilla NSGD.

We tune S-Muon with $\alpha=0.5$ for the best \texttt{(lr, momentum, sign\_lr\_coeff)} and by setting \texttt{lr=0.42(1 - step/total\_steps), momentum=0.63, nesterov = True, sign\_lr\_coeff = 0.003} get $\mathbf{94.03} \pm 0.13\%$, which is even higher than for vanilla Muon.

@@

The Muon-level performance is surprising for the contorted geometry of F-Muon and S-Muon. F-Muon can be visualized by \figureref{fig:cifar_ball}, where the LMO balls of Muon and F-Muon for real \texttt{lr} are plotted in a 2D space of singular values. S-Muon cannot be visualized because the $l_\infty$ norm is not a function of singular values, while $\norms{\cdot}$, when plotted for the $\Rmn$ space with $m, n > 2$, requires at least a 4D space. However, it is plain that the LMO ball of S-Muon significantly differs from the ball of Muon because the $\texttt{lr}(1-\alpha)\texttt{sign\_lr\_coeff} = 6.3 \times 10^{-4}$, which is comparable to the usual \texttt{lr} of SignSGD for deep learning problems.

The most pathological case is $\alpha > 1$, which corresponds to the LMO with an area that is not a norm ball! Despite this violation, the mixture of algorithms reaches almost the same accuracy as vanilla Muon.

These observations raise the question of how much LMO-based algorithms are sensitive to the constraint area.

@@

We test F-Muon and S-Muon with $\alpha=0.5$ on the NanoGPT speedrun \citep{modded_nanogpt_2024}. The optimal parameters are in the legend \figureref{fig:gpt_loss}, \texttt{sign\_sgd\_coeff=0.0003/0.07}. After testing for 1750 steps we get 3.281 cross-entropy loss for F-Muon and 3.287 for S-Muon. For Muon, it falls below the target threshold 3.28 and reaches 3.279. However, this difference is negligible if one looks at \figureref{fig:gpt_loss}. It is even more striking, considering the fact that this F-Muon is an average of Muon and NSGD, and the latter performed quite poorly. The same is true for S-Muon.

@@

\subsection{GPT-2 Medium Speedrun}
We scale from NanoGPT to GPT-2 Medium. After testing for 5960 steps with F-Muon we get 2.9215 loss, while for Muon it is 2.9198. In ? steps, F-Muon reaches <2.92 loss, thus needing only ?\% more iterations than Muon.

@@

Common important drawback of the analyses is the consideration of the convergence by the norm of the gradient. As we showed in our experiments on CIFAR, the gradient norm may decrease only by a factor of ten, when the accuracy reaches 100\%.

@@

\section{Under the hood of CNN on CIFAR-10}

>>>>

We then compare the performance of these algorithm families on various model and real-world problems:
\begin{itemize}
	\item Synthetic least squares experiment \sectionref{subsec:lls_exps}
	\item CIFAR-10 airbench \citep{cifar2023airbench}
	\item Pre-training NanoGPT and GPT-2 Medium on FineWeb dataset \citep{modded_nanogpt_2024}
	\item Fine-tuning NanoGPT on Tiny Stories \citep{eldan2023tinystoriessmalllanguagemodels}
\end{itemize}

@@

Training a neural network is essentially an optimization of a function of several weight matrices and a few vectors. Let us start by considering the problem of minimizing a differentiable function of a \emph{single} matrix:
\begin{equation}\label{eq:matrix_problem}
	F(\cdot)\colon \Rmn \to \R\,,\qquad F(\mX) \to \min_{\mX \in \Rmn}\,
\end{equation}
We equip the matrix space $\Rmn$ with a standard dot product $\<\cdot, \cdot> \to \R$ and a norm $\norm{\cdot}\colon \Rmn \to \R_+$, which does not have to coincide with the Frobenius norm $\normf{\cdot} = \<\cdot,\cdot>$. The dual norm $\norm{\cdot}^\dagger\colon \Rmn \to \R_+$ that is associated with $\norm{\cdot}$ is defined as
\begin{equation}\label{eq:dual_norm}
	\norm{\mM}^\dagger = \sup_{\mD \in \Rmn\,:\norm{\mD}\leq 1} \<\mM,\mD>\,.
\end{equation}

Such problems can be solved with an iterative algorithm based on the Linear Minimization Oracle (LMO):
\begin{equation}\label{eq:lmo}
	\mathrm{LMO}(\mM^t) \in \argmin_{\mD \in \cS} \langle\mM^t, \mD\rangle\,,
\end{equation}
where $\mM^t$ is the effective update direction and $\cS \subset \Rmn$ is a constraint set. The algorithm proceeds as:
\begin{equation}\label{eq:simple_update}
  \begin{aligned}
  \mB^t &= \beta \mB^{t-1} + (1 - \beta) \nabla F(\mX^t) \quad \text{(momentum buffer)},\\
  \mM^t &= \begin{cases}
  \nabla F(\mX^t) & \text{(no momentum)},\\
  \mB^t & \text{(heavy ball)},\\
  \nabla F(\mX^t) + \beta \mB^t & \text{(approximate Nesterov)},
  \end{cases}\\
  \mX^{t + 1} &= \mX^t + \gamma_t\mathrm{LMO}\left(\mM^t\right)\,,
  \end{aligned}
\end{equation}
where $\beta \in [0, 1)$ is the momentum coefficient. Throughout our experiments, we employ approximate Nesterov momentum, which matches the implementation in Muon and PyTorch SGD (\texttt{nesterov=True}). It uses the current gradient rather than the true lookahead gradient, trading theoretical guarantees for computational efficiency.

We are particularly interested in the case when $\cS$ is a unit ball in some norm $\norm{\cdot}$:
\[
\cS = \cB_{\norm{\cdot}} = \{ \mD \in \Rmn \mid \norm{\mD} \leq 1 \}\,.
\]
In this case,
\[
\argmin_{\mD \in \cS} \<\mM^t, \mD> = - \{ \mD \in \cB_{\norm{\cdot}} \mid \<\mM^t, \mD> = \norm{\mM^t}^\dagger\}\,,
\]
and the update for $\mX^{t + 1}$ in \equationref{eq:simple_update} simplifies to
\begin{equation}\label{eq:our_update}
	\mX^{t+1} = \mX^{t} - \gamma_t \{\mD \in \cB_{\norm{\cdot}} \mid \<\mM^t, \mD> = \norm{\mM^t}^\dagger\}\,.
\end{equation}

@@

\begin{corollary}\label{corollary:lin_comb_alg}
	Let there be a finite family of LMO based algorithms indexed by $i = 1,\dots,n$, where the $\mX^{t + 1}$ update in the $i$-th algorithm is defined by
	\[
	\mX^{t + 1} - \mX^{t} = \gamma_t \mathrm{LMO}_{i}(\mM^{t})\,,
	\]
	and $\mathrm{LMO}_i$ corresponds to the unit ball of norm $\norm{\cdot}_i$. For arbitrary non-negative $\alpha_1,\dots,\alpha_n$, the algorithm with the update given by
	\[
	\mX^{t + 1} - \mX^{t} = \gamma_t \sum_{i = 1}^n \alpha_i \mathrm{LMO}_{\norm{\cdot}_{(i)}}(\mM^t)
	\]
	is an LMO-algorithm itself, with LMO corresponding to the unit ball of the norm $\norm{\cdot}$ dual to the norm given by $\sum_{i = 1}^n \alpha_i \norm{\cdot}_{(i)}^\dagger$.
\end{corollary}

@@

Similarly, combining Fanion-$k$ with SignSGD yields \emph{S-Fanion-$k$} with update
\begin{equation}\label{eq:sign_muon_update}
		\mX^{t+1} = \mX^{t} - \gamma_t \left(\alpha \sum_{i = 1}^k \vu_i \vv_i^\top + (1-\alpha)\eta\sign(\mM^t)\right)\,,
\end{equation}
where \texttt{sign\_lr\_coeff} $\eta$ is a scaling coefficient specific to SignSGD.

@@

We begin by evaluating F-Fanions on the following convex $L$-smooth problem:
\begin{equation}\label{eq:lls}
  F(\mX) = \frac{1}{2} \<(\mX-\mS), \mM (\mX - \mS) \mN> \to \min_{\mX \in \Rmn}\,
\end{equation}
where $\mX \in \Rmn$, $m=500$ and $n=500$ represent typical dimensions of a neural network weight matrix, $\mS \in \Rmn$, and $\mM \in \mathbb{S}^m_{+}$ and $\mN \in \mathbb{S}^n_{+}$ are positive-semidefinite matrices. The spectra of $\mM$ and $\mN$ are uniformly distributed on the interval $(0, 1)$. We set $\mS=0$ and initialize $\mX^0$ with entries drawn independently from $\cN(0, 0.01)$.

We evaluate several algorithms from the Fanion family and their convex combinations, using the update rule from \equationref{eq:simple_update} with approximate Nesterov momentum:
\begin{itemize}
  \item \textbf{Fanions:} Neon (Fanion-1), Fanion-2, Fanion-10, Fanion-100, and Muon (Fanion-500).
  \item \textbf{F-Fanions:} F-Neon (F-Fanion-1), F-Fanion-2, F-Fanion-10, F-Fanion-100, and F-Muon (F-Fanion-500) with $\alpha=1/2$.
  \item \textbf{S-Fanions:} S-Neon (S-Fanion-1), S-Fanion-2, S-Fanion-10, S-Fanion-100, and S-Muon (S-Fanion-500) with $\alpha=1/2$ and \texttt{sign\_lr\_coeff=0.01}.
\end{itemize}
We also include baselines Normalized SGD and SignSGD, which correspond to F-Fanion and S-Fanion respectively with $\alpha = 0$ and arbitrary $k$.

Since theoretical bounds \citep{kovalev2025understanding, riabinin2025gluon} rely on a loose norm bound $\norm{\cdot}\le \rho \normf{\cdot}$, we do not derive the learning rate or Nesterov momentum from the smoothness constants, which also depend on the choice of norm. Instead, we identify the \texttt{(lr, momentum)} pair that reaches the loss threshold of $0.001$ in the minimal number of iterations. This setting is both realistic and consistent with the corollaries of convergence theorems that propose constant learning rate and momentum coefficients.

We perform a grid search over the following hyperparameter ranges:
\begin{itemize}
  \item \texttt{momentum} $\in \{0.1, 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 0.95\}$ for all algorithms.
  \item \texttt{lr} $\in [0.005, 0.020]$ with step $0.001$ for Muon, F-Muon, and S-Muon.
  \item \texttt{lr} $\in [5\mathrm{e}{-5},\, 2\mathrm{e}{-4}]$ with step $1\mathrm{e}{-5}$.
  \item \texttt{lr} $\in [0.01, 0.10]$ with step $0.001$ for NSGD.
\end{itemize}
The tuned parameters and the number of iterations required to converge to $0.001$ are presented in \tableref{tbl:lls_lrs}. For intermediate-rank Fanions, F-Fanions, and S-Fanions, the hyperparameters \texttt{lr}, \texttt{momentum}, and \texttt{sign\_lr\_coeff} are set equal to those of Muon, F-Muon, and S-Muon respectively, as their loss decreases too slowly to reach $0.001$ within a reasonable tuning budget.

@@

The results are presented in \figureref{fig:lls} with additional details in \sectionref{sec:lls_plots_section}.

@@

Both F-Muon and S-Muon converge faster to lower loss values and achieve lower Frobenius norms of the gradient than NSGD, Muon, or SignSGD.

@@

We evaluate the algorithms on the CIFAR-10 airbench \citep{cifar2023airbench}. To assess the impact of the mixing parameter $\alpha$, we first run F-Muon for different values of $\alpha$ using hyperparameters tuned for vanilla Muon by Keller Jordan: \texttt{lr=0.24(1 - step/total\_steps), momentum=0.65, nesterov=True} with weight normalization. We perform 10 repetitions for each $\alpha$ value and record the accuracy after 8 epochs of training (\figureref{fig:muon_alphas}).

Next, we tune F-Muon specifically with $\alpha = 0.5$, obtaining optimal hyperparameters \texttt{lr=0.4(1 - step/total\_steps), momentum=0.6, nesterov = True}. Tuned F-Muon achieves $94.02 \pm 0.13\%$ validation accuracy after 8 epochs (averaged over 200 runs), matching Muon's accuracy and variance. We again measure the validation accuracy as a function of $\alpha$ (\figureref{fig:fmuon_alphas}) and observe that even at $\alpha=0.1$, the accuracy substantially exceeds that of vanilla NSGD.

For S-Muon with $\alpha=0.5$, we tune all hyperparameters to find the optimal configuration \texttt{lr=0.42(1 - step/total\_steps), momentum=0.63, nesterov = True, sign\_lr\_coeff = 0.003}, achieving $\mathbf{94.03} \pm 0.13\%$ validation accuracy, which slightly exceeds vanilla Muon's performance.

@@

The Muon-level performance is remarkable given that F-Muon and S-Muon operate with substantially different constraint geometries. \figureref{fig:cifar_ball} visualizes F-Muon by plotting the LMO balls of Muon and F-Muon (scaled by actual learning rates) in the 2D space of singular values. S-Muon cannot be visualized in this manner because the $l_\infty$ norm is not a function of singular values alone, and visualizing $\norms{\cdot}$ for $\Rmn$ with $m, n > 2$ requires at least 4D space. Nevertheless, the LMO ball of S-Muon clearly differs substantially from Muon's, as the effective SignSGD learning rate $\texttt{lr}(1-\alpha)\texttt{sign\_lr\_coeff} = 6.3 \times 10^{-4}$ is comparable to typical SignSGD learning rates in deep learning.

Notably, even the pathological case $\alpha > 1$—which corresponds to an LMO constraint region that is not a norm ball—achieves accuracy nearly identical to vanilla Muon.

These observations raise fundamental questions about the sensitivity of LMO-based algorithms to the shape of the constraint region.

@@

We evaluate F-Muon and S-Muon with $\alpha=0.5$ against Muon, SignSGD, and NSGD on the NanoGPT speedrun benchmark \citep{modded_nanogpt_2024}. The optimal hyperparameters are shown in the legend of \figureref{fig:gpt_loss}, with \texttt{sign\_sgd\_coeff} values of 0.0003/0.07 for S-Muon. After 1750 training steps, F-Muon achieves a cross-entropy loss of 3.281, S-Muon achieves 3.287, and Muon achieves 3.279, falling below the target threshold of 3.280. As illustrated in \figureref{fig:gpt_loss}, these differences are negligible. Notably, F-Muon represents a convex combination of Muon and NSGD, despite NSGD performing poorly in isolation. The same observation holds for S-Muon.

@@

\subsection{GPT-2 Medium Speedrun}
We scale from NanoGPT to GPT-2 Medium. After testing for 5960 steps we get 2.9215 loss for F-Muon and 2.9235 for S-Muon, while for Muon it is 2.9198 (the speedrun threshold is 2.92).

@@

Common important drawback of the analyses is the consideration of the convergence by the norm of the gradient. As we showed in our experiments on CIFAR \cref{sec:cifar_underhood}, the gradient norm may decrease only by a factor of ten even when the accuracy reaches 100\%.

@@

\section{Under the hood of CNN on CIFAR-10}
\label{sec:cifar_underhood}
